QPLIB: A Library of Quadratic Programming Instances

### 2017-01-27 QPLIB-3.tex
Add files via upload

For each instance of the starting set, we collected important characteristics
which allowed us to classify the instances into the QP categories described in
\S~\ref{sec:QPbasic}. As far as the variable types are concerned, we
collected the following information:
%
\begin{itemize}
 \item number of binary variables; % (\emph{\# bin}),
 %
 \item number of integer variables; % (\emph{\# int}),
 %
 \item number of continuous variables. % (\emph{\# cont}).
\end{itemize}
%
If at least one binary or integer variable is present, then the instance is
categorized as \emph{discrete}, otherwise it is categorized as \emph{continuous}.
As far as the objective function is concerned, we gathered the following
information:
%
\begin{itemize}
 \item percentage of negative eigenvalues of the $Q^0$ matrix;
       % (\emph{\% neg eig}),
 %
 \item density of the $Q^0$ matrix (number of nonzero entries divided by the total
       number of entries). %(\emph{\% dens}).
\end{itemize}
%
The number of negative eigenvalues of $Q^0$ allows us to identify the
objective function type, as in presence of at least one negative eigenvalue
the objective function is non convex. Finally, as far as the constraint types
are concerned, we collected the following information:
%
\begin{itemize}
 \item number of linear constraints, % (\emph{\# lin}),
 %
 \item number of quadratic constraints, % (\emph{\# quad}),
 %
 \item number of convex constraints, % (\emph{\# conv}),
 %
 \item number of box constraints. %(\emph{\# box}).
\end{itemize}
%
A constraint is considered quadratic if it contains at least one nonzero in
the quadratic term. Among the quadratic constraints, the ones with only
non-negative eigenvalues are classified as convex constraints. All this
information allowed us to analyse the gathered instances and to perform the
filters described in the the next paragraph.

@@

During the development of the library, a discussion ensued about
the expected goals that we wanted to achieve. The following four goals
where finally identified:
%
\begin{enumerate}
 \item represent as much as possible all the different categories of QP
       problems;
 %
 \item gather ``challenging'' instances, i.e., ones which can not be easily
       solved by  state-of-the-art solvers;
 %
 \item include for each of the categories a set of well-diversified
       instances;
 %
 \item obtain a set of instances which is neither too small, so as to
       preserve statistical relevance, nor too large so as to being
       computationally manageable.
\end{enumerate}
%
To achieve the aforementioned goals, we performed the following two
filters, applied in cascade.
%
\begin{itemize}
 \item \emph{First Instances Filter.}\\
       The first filter was designed to drastically reduce the number of
       instances by eliminating the ``easy'' ones. An empirical measure
       for the hardness of an instance is the CPU time needed by a
       complete solver (cf.~\S~\ref{sec:algo}) to solve it to
       global optimality. Accordingly, for each of the gathered instance we
       ran the complete solvers in GAMS, which number depends on the category
       of the instance under consideration, cf.~Table \ref{t:solvers}.
       We then filtered according to a relative
       measure of computational difficulty, i.e., we discarded all instances
       that are solved by at least 30\% of the complete solvers within a time
       limit of 30 seconds.
 %
 \item \emph{Second Instances Filter.}\\
       The goal of the second filter was to eliminate ``similar'' instances.
       We carefully analysed the instances one by one, and we clustered them
       according to their features; for each cluster we kept only a few
       representatives\todo{we may want to be a tad more specific
       about what ``features'' exactly meant there}. Finally, in order to only
       keep computationally challenging instances we ran a general purpose solver
       able to tackle all QP categories\todo{do you mean a complete solver for QGQ? if so, maybe say so}\xspace with a time limit of 120 seconds; all the
       instances which have been solved to proven optimality within this time limit
       were discarded\todo{this might actually be best characterized as
       a third filter}.
\end{itemize}
%
In Table~\ref{tab:filters} we summarize the two filter steps, which
allowed us to identify the final set of $251$ discrete instances and
$116$ continuous instances.

@@

We now analyse the features of the instances selected to be part of the
library. The characteristics of the instances are presented in Table
\ref{tab:DD} for \emph{discrete} instances (*\{B,M,I,G\}*) and in Table
\ref{tab:CC} for \emph{continuous} ones (*C*). For each category, the tables
report in column ``$\#$'' the corresponding number of instances. It can be seen
that the final set well respects the original distribution of the gathered
instances\todo{Maybe one could put the numbers for the original set of instances
into Tables \ref{tab:DD} and \ref{tab:CC}, too?}\xspace among the different categories. Indeed, the discrete categories
(LMQ) or (QBL) are well represented by $118$ and $59$ instances,
respectively. Similarly, the continuous categories (LCQ) and (QCQ) are well
represented by $50$ and $17$ instances, respectively. Moreover, the library
actually covers the large majority of all possible categories of instances.

@@

Figure~\ref{fig:pic_conv_constr} shows the instances with at least one quadratic constraint sorted according to the number of quadratic constraints (vertical axis). For each instance we report the total number of constraints with a ``$+$''
and the total number of nonconvex quadratic constraints
with a ``$\times$''.

\begin{figure}\centering
  \includegraphics[width=0.85\textwidth]{pic_quad_conv_vs_nonconv.png}
  \todo[inline]{horizontal axis label could say ``Instances with at least one quadratic constraint''}
  \caption{Quadratic constraints \label{fig:pic_conv_constr}}
\end{figure}

Speaking of nonconvexity, Figure~\ref{fig:pic_neg_eig} focuses on
the instances with a quadratic objective function, ordered by the
the percentage of negative eigenvalues\todo{Was there some tolerance applied on when an eigenvalue is considered negative? $-10^{-9}$, $-10^{-6}$, etc.?} in the $Q^0$ matrix (vertical axis). The instances are mostly
clustered around two values. About 25\% of the instances are convex\todo{so one also restrict to linear constraints, or may only the objective be convex?},
i.e., they have 0\% of negative eigenvalues. Among the others, a vast
majority has around 50\% of negative eigenvalues. However, instances
with high or low percentages of negative eigenvalues are present, too.

\begin{figure}\centering
  \includegraphics[width=0.85\textwidth]{pic_neg_eig.png}
  \todo[inline]{horizontal axis label should say ``Instances with quadratic objective function''}
  \caption{Negative eigenvalues \label{fig:pic_neg_eig}}
\end{figure}

@@

\begin{figure}\centering
  \includegraphics[width=0.85\textwidth]{pic_density.png}
  \todo[inline]{horizontal axis label should say ``Instances with quadratic objective function''}
  \caption{Density \label{fig:pic_density}}
\end{figure}

Additional details on the instance features can be found in \S~\ref{sec:instance_details}.
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
%  End QPLIB-3.tex
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

>>>>

For each instance of the starting set, we collected important characteristics
which allowed us to classify the instances into the QP categories described in
\S~\ref{sec:QPbasic}. As far as the variable types are concerned, we
collected the following information:
%
\begin{itemize}
 \item number of binary variables; % (\emph{\# bin}),
 %
 \item number of integer variables; % (\emph{\# int}),
 %
 \item number of continuous variables. % (\emph{\# cont}).
\end{itemize}
%
If at least one binary or integer variable is present, then the instance is
categorized as \emph{discrete}, otherwise it is categorized as \emph{continuous}.
As far as the objective function is concerned, we gathered the following
information:
%
\begin{itemize}
 \item percentage of negative eigenvalues of the $Q^0$ matrix;
       % (\emph{\% neg eig}),
 %
 \item density of the $Q^0$ matrix (number of nonzero entries divided by the total
       number of entries). %(\emph{\% dens}).
\end{itemize}
%

The number of negative (i.e., smaller than $-10^{-16}$) eigenvalues of $Q^0$ allows us to identify the
objective function type, as in presence of at least one negative eigenvalue
the objective function is non convex. Finally, as far as the constraint types
are concerned, we collected the following information:
%
\begin{itemize}
 \item number of linear constraints, % (\emph{\# lin}),
 %
 \item number of quadratic constraints, % (\emph{\# quad}),
 %
 \item number of convex constraints, % (\emph{\# conv}),
 %
 \item number of box constraints. %(\emph{\# box}).
\end{itemize}
%
A constraint is considered quadratic if it contains at least one nonzero in
the quadratic term. Among the quadratic constraints, the ones with only
non-negative eigenvalues are classified as convex constraints.
Note that this might lead to classify some instances that have conic constraints as non-convex ones, although their feasible region is in fact convex. However, only some solvers are capable of properly exploiting this property.
 All this information allowed us to analyse the gathered instances and to perform the
filters described in the the next paragraph.

@@

During the development of the library, a discussion ensued about
the expected goals that we wanted to achieve. The following four goals
where finally identified:
%
\begin{enumerate}
 \item represent as much as possible all the different categories of QP
       problems;
 %
 \item gather ``challenging'' instances, i.e., ones which can not be easily
       solved by  state-of-the-art solvers;
 %
 \item include for each of the categories a set of well-diversified
       instances;
 %
 \item obtain a set of instances which is neither too small, so as to
       preserve statistical relevance, nor too large so as to being
       computationally manageable.
\end{enumerate}
%
To achieve the aforementioned goals, we performed the following two
filters, applied in cascade.
%
\begin{itemize}
 \item \emph{First Instances Filter.}\\
       The first filter was designed to drastically reduce the number of
       instances by eliminating the ``easy'' ones. An empirical measure
       for the hardness of an instance is the CPU time needed by a
       complete solver (cf.~\S~\ref{sec:algo}) to solve it to
       global optimality. Accordingly, for each of the gathered instance we
       ran the complete solvers in GAMS, which number depends on the category
       of the instance under consideration, cf.~Table \ref{t:solvers}.
       We then filtered according to a first
       measure of computational difficulty, i.e., we discarded all instances
       that are solved by at least 30\% of the complete solvers within a time
       limit of 30 seconds.
 %
 \item \emph{Second Instances Filter.}\\
       The goal of the second filter was to eliminate ``similar'' instances.
       We carefully analysed the instances one by one, and we clustered them
       according to their features; for each cluster we kept only a few
       representatives (e.g., very similar size, same donor,\ldots). Finally, in order to only
       keep computationally challenging instances we ran a a complete solver for QGQ with a time limit of 120 seconds; all the
       instances which have been solved to proven optimality within this time limit
       were discarded.
\end{itemize}
%
In Table~\ref{tab:filters} we summarize the two filter steps, which
allowed us to identify the final set of $251$ discrete instances and
$116$ continuous instances.

@@

We now analyse the features of the instances selected to be part of the
library. The characteristics of the instances are presented in Table
\ref{tab:DD} for \emph{discrete} instances (*\{B,M,I,G\}*) and in Table
\ref{tab:CC} for \emph{continuous} ones (*C*). For each category, the tables
report in column ``$\#$'' the corresponding number of instances. It can be seen
that the final set well respects the original distribution of the gathered
instances among the different categories. Indeed, the discrete categories
(LMQ) or (QBL) are well represented by $118$ and $59$ instances,
respectively. Similarly, the continuous categories (LCQ) and (QCQ) are well
represented by $50$ and $17$ instances, respectively. Moreover, the library
actually covers the large majority of all possible categories of instances.

@@

Figure~\ref{fig:pic_conv_constr} shows the instances with at least one quadratic constraint sorted according to the number of quadratic constraints (vertical axis). For each instance we report the total number of constraints with a ``$+$''
and the total number of nonconvex quadratic constraints
with a ``$\times$''.
It can be seen that the majority of instances only have convex constraints; of those that have nonconvex ones, however, a significant fraction have no convex ones at all.

\begin{figure}\centering
  \includegraphics[width=0.85\textwidth]{pic_quad_conv_vs_nonconv.png}
  \caption{Quadratic constraints (instances with at least one) \label{fig:pic_conv_constr}}
\end{figure}

Speaking of nonconvexity, Figure~\ref{fig:pic_neg_eig} focuses on
the instances with a quadratic objective function, ordered by the
the percentage of negative eigenvalues in the $Q^0$ matrix (vertical axis). The instances are mostly
clustered around two values. About 25\% of the instances have convex objective function,
i.e., they have 0\% of negative eigenvalues. Among the others, a vast
majority has around 50\% of negative eigenvalues. However, instances
with high or low percentages of negative eigenvalues are present, too.

\begin{figure}\centering
  \includegraphics[width=0.85\textwidth]{pic_neg_eig.png}
  \caption{Negative eigenvalues (instances with quadratic objective function) \label{fig:pic_neg_eig}}
\end{figure}

@@

\begin{figure}\centering
  \includegraphics[width=0.85\textwidth]{pic_density.png}
  \caption{Density of $Q^0$ (instances with quadratic objective function) \label{fig:pic_density}}
\end{figure}

Additional details on the instance features can be found in \S~\ref{sec:instance_details}.

%The all set of instances can be downloaded at \url{http://qplib.zib.de/html/index.html}.
%Since, as described in Section \ref{subsec:final set} the library contains instances of different nature, the website allows to filter the instances according to the classification proposed in Section \ref{ssec:taxonomy}.

\subsection{Website}
%Select subset or categories of instances (EXTRACT FROM THE LIBRARY A SUBSET OF INSTANCES WITH SPECIFIC CHARACTERISTICS)
%\todo[inline]{TASK X : write }

The instances of QPLIB are publically accessible at the website \url{http://qplib.zib.de}.
%
Beyond links to the \texttt{gms} and \texttt{lp} format files, the website provides a rich set of metadata for each
instance: the three letter problem classification (as described in Section \ref{subsec:final set}), basic properties such as the number of variables and constraints of
different types, the sense and curvature of the objective function, and information on the nonzero structure of the
problem.
%
In addition, we display a visualization of the sparsity patterns of the Jacobian and the Hessian matrix of the
Lagrangian function.  In the plots of the Jacobian nonzero pattern, the linear and nonlinear entries are distinguished
by color.  Figure... shows an example for instance QPLIB...

The entire set of instances can be explored in a searchable and sortable table of selected instance features: problem
classification, convexity of the continuous relaxation, and number of variables, constraints, and nonzeros.
%
Finally, a statistics page displays diagrams on the composition of the library according to different criteria: the
number instances according to problem type, variable types, convexity, number of variables and constraints, and density.
%
A table containing the relevant metadata for each instance can be downloaded in \texttt{csv} format and as spreadsheets
such that researchers can easily compile their own subset of instances according to these statistics.

The complete library can be downloaded as one archive, which contains the website for offline browsing and exploration.  
%
In the future, we plan to extend the website by the addition of contributor information and references to the literature.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
%  End QPLIB-3.tex
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
